\texttt{aeons}: approximating the end of nested sampling

### 2023-11-10 paper/paper.tex
Further updates to text

\section{Background}\label{sec:background}
We begin with a brief description of the nested sampling algorithm to establish the necessary notation. For a more comprehensive treatment, we recommend the original \citep{skilling} paper, the \citet{sivia} textbook, as well as the excellent technical review by \citet{Buchner_2023} and Nature review by \citet{physical_scientists}. 
\par
For a given likelihood $\mathcal{L}(\theta)$ and prior $\pi(\theta)$, nested sampling simultaneously calculates the Bayesian evidence
\begin{equation}\label{eq:evidence}
	\mathcal{Z} = \int \mathcal{L}\left(\theta\right)\pi(\theta)\ \mathrm{d}\theta
\end{equation}
while producing samples of the posterior distribution
\begin{equation}
	\mathcal{P}(\theta) = \frac{\mathcal{L}(\theta) \pi(\theta)}{\mathcal{Z}}.
\end{equation}
The algorithm operates by maintaining a set of $\nlive$ \textit{live points} sampled from the prior, which can vary in number throughout the run \citep{dynamic_ns}. At each iteration $i$, the point with the lowest likelihood is removed and added to a list of \textit{dead points}. A new point is then drawn from the prior, subject to the constraint that it must have a higher likelihood than the latest dead point. Repeating the procedure leads to the live points shrinking around peaks in the likelihood.
\par
The integral in \eqref{eq:evidence} is then evaluated by transformation to a one-dimensional integral over the \textit{prior volume} $X$
\begin{equation}
	\mathcal{Z} = \int_0^1 \mathcal{L}(X)\ \mathrm{d}X \approx \frac{1}{2}\sum_{i=1} \mathcal{L}(X_{i-1}-X_{i+1}),
\end{equation}
where $X(\mathcal{L})$ is the fraction of the prior with a likelihood greater than $\mathcal{L}$. The prior volumes $X_i$ are unknown, but can be statistically estimated as follows: one can define a \textit{shrinkage factor} $t_i$ at each iteration $X_{i} = t_i X_{i-1}$, such that
\begin{equation}\label{eq:X_dist}
	X_i = \prod_{k=1}^i t_k.
\end{equation}
The $t_i$ are the maximum of $\nlive$ points drawn from $[0,1]$, so follow the distribution
\begin{equation}\label{eq:t_dist}
	P(t_i) = \nlive t_i^{\nlive-1},
\end{equation}
\begin{equation}\label{eq:t_moments}
    \langle\log t_i\rangle = -\frac{1}{\nlive}, \quad \mathrm{Var}(\log t_i) = \frac{1}{\nlive^2}.
\end{equation}
The algorithm terminates when an user-specified condition is met; a popular choice is when the evidence in the live points falls below some fraction $\epsilon$ of the accumulated evidence e.g. $10^{-3}$, which is proven to be a valid convergence criterion \citep{evans, Chopin_2010}.
Much of the existing literature treats this remaining evidence separately, for instance by estimating it as the termination $X$ multiplied by the average likelihood amongst the remaining live points. It is, however, quantitatively equivalent but qualitatively neater to consider termination as killing the remaining live points off one-by-one, incrementing the evidence exactly as during the run with decreasing $\nlive$ \citep{dynesty}.
\par
Uncertainties in the evidence are dominated by the spread in the prior volume distribution, and the simplest way to estimate them is by Monte Carlo sampling over sets of $\bm{t}$. As Skilling and others \citep{Chopin_2010, Keeton_2011} have shown, for any given problem the uncertainty in $\log \mathcal{Z}$ is proportional to $1/\sqrt{n_\mathrm{live}}$, so $n_\mathrm{live}$ sets the resolution of the algorithm.

\section{The anatomy of a nested sampling run}\label{sec:anatomy}
The following section aims to make an inventory of the information available to us at an intermediate iteration $i$, which we shall eventually use to make endpoint predictions. We present an anatomy of the progression of a nested sampling run in terms of the prior volume compression, the log-likelihood increase, the inferred temperature, and the dimensionality of the samples.

\subsection{Prior volume}
The key feature of nested sampling is that the sampling is controlled by prior volume compression. The task is to find the posterior typically lying in a tiny fraction of the prior volume, a total compression which is quantified by the average information gain, or \textit{Kullback-Leibler divergence}:
\begin{equation}\label{eq:DKL}
   \DKL = \int \mathcal{P}(\theta) \log \frac{\mathcal{P}(\theta)}{\pi(\theta)}\ \mathrm{d}\theta. 
\end{equation}
The bulk of the posterior lies within a prior volume $X = e^{-\DKL}$, which is the target compression. One gets there by iteratively taking steps of size $\Delta \log X_i = -1/n_i$, so that when we add up the contribution of each step in \eqref{eq:t_moments} we get
\begin{equation}
    \langle\log X_i\rangle = -\sum_{k=1}^i \frac{1}{n_k}, \quad \mathrm{Var}(\log X_i) = \sum_{k=1}^i \frac{1}{n_k^2}.
\end{equation}
A constant step size in $\log X$ corresponds to a geometrically constant measure for the dead points (as shown in \cref{fig:dead_measure}), which is exactly needed to overcome the curse of dimensionality.
\begin{figure*}
\begin{center}
    \includegraphics{figures/dead_measure_live.pdf}
\end{center}
\caption{The dead points of a nested sampling run, recursively zoomed in. Their density is constant in $\log X$, which is a geometrically constant measure $\propto \mathrm{d} V/V$, and hence scale invariant. The live points (larger dots) are uniform across the prior, plotted for comparison.}
\label{fig:dead_measure}
\end{figure*}
\par
The same is not true for the live points, which are uniformly distributed in prior volume. As a result, the maximum live point is found at 
\begin{equation}\label{eq:Xmin}
	\langle\log X_\mathrm{min}^{\mathrm{live}}\rangle = \langle\log X_i\rangle - \sum_{k=1}^{\nlive} \frac{1}{n_k} \approx -\frac{i}{\nlive} - \log \nlive - \gamma,
\end{equation}
with variance
\begin{equation}
	\mathrm{Var}(\log X_\mathrm{min}^{\mathrm{live}}) = \mathrm{Var}(\log X_i) + \sum_{k=1}^{\nlive} \frac{1}{n_k^2} \approx \frac{i}{\nlive^2} + \frac{\pi^2}{6}, 
\end{equation}
where the large $\nlive$ limit is taken for the approximation to the harmonic series, $\gamma$ being the Euler-Mascheroni constant.
\par
The live points therefore only get us a factor of $\log \nlive$ closer to the posterior bulk. In other words, it is not until we are around $\log \nlive$ away from $\log X = -\DKL$ that the samples look anything like the posterior. One can see from \eqref{eq:DKL} that the divergence increases linearly with dimension, so for large dimensionalities and typical live point numbers $\lesssim 1000$, this does not happen until near the end of the run.
\par
The result is consistent with that in \citet{statmech}, which states that a spike at a volume smaller than $X_i/n_i$ will go undetected. Intuitively, it is because for a sharply peaked likelihood the live points are too diffuse to land there with any significant probability for most of the run.
\par
These results are summarised in \cref{fig:logX_distribution}
\begin{figure*}
\begin{center}
    \includegraphics{figures/logX_distribution.pdf}
\end{center}
\caption{The distribution of the posterior mass in terms of $\log X$, the live points over the constrained prior and the smallest live point prior volume $\log X_\mathrm{min}^{\mathrm{live}}$ at an intermediate iteration $i$. For large values of $\DKL$ i.e. informative posteriors and/or large dimensionalities, the maximum live point is very far from the posterior bulk until the very end of the run. Note that the x-axis in this plot is $\log X$, so that the run proceeds from right to left to emphasise that the enclosed prior volume iteratively gets smaller. Plots of the sort from here onward will be in terms of $-\log X$, where the run will more naturally proceed from left to right. }
\label{fig:logX_distribution}
\end{figure*}

\subsection{Log-likelihood}\label{sec:logL}
It is also useful to observe the distribution of the live and dead points in log-likelihood. We choose to examine a representative case of the $d$-dimensional multivariate Gaussian:
\begin{equation}\label{eq:gaussian_logL}
    \log \Like = \logLmax - X^{2/d}/2\sigma^2
\end{equation}
with maximum point $\logLmax$ and length scale $\sigma$ to get an insight into the analytics. The likelihood monotonically increases towards $\logLmax$, with the posterior samples eventually concentrated around the bulk at
\begin{equation}
    \langle\log\Like\rangle_\mathcal{P} = \log\mathcal{L}_\mathrm{max} - \frac{d}{2},  \quad \mathrm{Var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}.
\end{equation}
We can again find the size of each step in log-likelihood, as well as the expected location of the maximum live point. Let us define a likelihood normalised by the distance to the maximum:
\begin{equation}
    y = \frac{\log\mathcal{L}-\log\mathcal{L}_i}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_i};
    \label{eq:normalised_likelihood}
\end{equation}
$y=0$ corresponds to the current point $\log\mathcal{L}_i$ and $y=1$ to the maximum $\log\mathcal{L}_\mathrm{max}$. At each iteration, $y$ increases by roughly
\begin{equation}
    \lim_{\nlive \to \infty}\Delta y \approx \frac{\mathrm{d} y}{\mathrm{d}\log X} \Delta \log X = \frac{2}{d \nlive}  
\end{equation}
in the large $\nlive$ limit. We can again sum the harmonic series to see that the maximum live point is expected to be at 
\begin{equation}
    y_{(\nlive)}^{\mathrm{live}} = \frac{2\log \nlive}{d},
\end{equation}
where we have used the notation of order statistics to denote $x_{(n)}$ as the maximum of $n$ points. This is a very small fraction in high dimensions, showing again that until the end the live points are far from the posterior bulk, and certainly nowhere near the maximum.
\par
The above also implies that the normalised distance between the highest and second highest live point is roughly 
\begin{equation}
    y_{(\nlive)}^{\mathrm{live}} - y_{(\nlive - 1)}^{\mathrm{live}} \approx \frac{2}{d}.
\end{equation}
Before reaching the posterior bulk, $\logLmax - \log\Like_i > d/2$, so we must have
\begin{equation}
    \log \Like^{\mathrm{live}}_{(\nlive)} - \log \Like^\mathrm{live}_{(\nlive - 1)} > 1.
\end{equation}
In other words, the highest likelihood point is always at least an order of magnitude greater than the second highest. It is therefore typically the case that nearly all of the posterior mass is concentrated in a single point, the maximum live point, until the very end of the run when the prior volumes have shrunk enough to compensate.

@@

\subsection{Temperature}
\subsubsection*{Motivations}
As shown in the previous section, midway through the run nearly all of the posterior mass is concentrated at a single point. However, this does not capture the \textit{structure} of the posterior that has been explored and all of the information it provides. 
\par
We have the potential to fix this because nested sampling is invariant to monotonic transformations, so we can transform the likelihood as $\Like \to \Like^{\beta}$ without loss of information by trivially re-weighting the samples. Increasing $\beta$ worsens the situation, while $\beta \to 0$ simply gives back the prior. There is, on the other hand, a significant intermediate range which makes the samples look like a posterior centred at the present contour, which will allow us to recover the structure of the samples. A schematic of the procedure is shown in \cref{fig:last_live_point}.
\par
\begin{figure}
\begin{center}
    \includegraphics{figures/last_live_point.pdf}
\end{center}
\caption{The likelihood and posterior as a function of $\log X$ in the middle of a nested sampling run. Almost all of the posterior mass is concentrated at a single point with the highest compression, because it is orders of magnitude higher in likelihood. Reducing $\beta$ re-weights to shift the posterior mass; taking $\beta \to 0$ goes too far and gives back the prior, but there is some intermediate $\beta = \beta_i$ which is significant because it centres the posterior at the current contour.}
\label{fig:last_live_point}
\end{figure}
\par
At this point it is relevant to note the correspondence between Bayesian inference and statistical mechanics, from which the above transform is derived. If one equates the parameters to microstates $i$, the negative log-likelihood to the microstate energy $E_i$, and the prior to the density of states $g_i$, then the posterior as given by the generalised Bayes' rule is the canonical ensemble
\begin{equation}
    p(E_i) = \frac{g_i e^{-\beta E_i}}{Z(\beta)} \quad \leftrightarrow \quad \mathcal{P}_\beta(\theta) = \frac{\mathcal{L}^{\beta}(\theta)\pi(\theta)}{\mathcal{Z}(\beta)}
\end{equation}
at the inverse temperature $\beta = 1/T$. As noted by \citet{demons}, thermal algorithms such as thermodynamic integration \citep{path_sampling} get the evidence by evolving through a series of canonical ensembles via some temperature schedule, but nested sampling instead maintains \textit{microcanonical} ensembles, which are the iso-likelihood contours. Instead of using temperature \citep{simulated_annealing, parallel_tempering} or energy \citep{wang_landau} as a control parameter, nested sampling chooses a series of ensembles with constant relative volume entropy $\Delta \log X$, which allows the algorithm to handle phase transitions \citep{baldock}.
\par
Because the temperature of a microcanonical ensemble is a derived property rather than a parameter, there is some freedom in its definition. Returning to our original motivation, we make the connection that the temperature is the re-weighting $\Like \to \Like^{\beta}$ which centres the ensemble around the current energy. We now present several temperatures that achieve this aim, each of which one can plausibly consider to be the current temperature of a nested sampling run.

\subsubsection{Microcanonical temperature}
The obvious candidate is the microcanonical temperature $\partial S/\partial E$, where the volume entropy is $\log X$ and the energy is as usual $-\log \Like$. This gives the density of states; as discussed in Skilling's original paper,
\begin{equation}
    \beta_\mathrm{M}  = - \frac{\mathrm{d} \log X}{\mathrm{d} \log \Like} \Bigg\vert_{\log \Like_i}.
\end{equation}
is the $\beta$ at which $\Like^{\beta} X$ peaks at $\log X_i$, if we assume differentiability, which is exactly the intuition we were aiming for to put the ensemble bulk at the current contour. 
\par
Its value can be easily obtained via finite difference of the $\log \Like$ and $\log X$ intervals, albeit subject to an arbitrary window size for the differencing. Indeed, material science applications \citep{Baldock_2017} use this estimator to monitor the `cooling' progress of nested sampling, with a window size of 1000 iterations.
\par
The ensemble bulk can be placed at the current point in several other plausible ways, which are discussed below.

\subsubsection{Canonical temperature}
Another temperature considered by Habeck is that at which the current energy (i.e. $-\log \Like_i$) is the average energy of the entire ensemble. One can obtain it by inverting 
\begin{equation}
    \langle \log \Like \rangle_{\mathcal{P}_\beta} = \log \Like_i
\end{equation}
to get the `canonical' temperature $\beta_\mathrm{C}$. While $\beta_\mathrm{M}$ is derived from (the gradient of) a single contour, this temperature uses the entire ensemble. It has the desirable property that it rises monotonically with compression, in analogy to a monotonic annealing schedule.

\subsubsection{Bayesian temperature}
We furthermore propose a temperature $\beta_\mathrm{B}$ that is obtained via Bayesian inference, which returns a distribution rather than a point estimate. Since each value of $\beta$ leads to a different likelihood $\Like^{\beta}$, one can consider the posterior distribution as a function of $\log X$ to be \textit{conditioned} on $\beta$. We can therefore write
\begin{equation}
   \mathcal{P}(\log X \mid \beta) = \frac{\Like^{\beta}(X) X}{\mathcal{Z(\beta)}}.
\end{equation}
What we would really like is the distribution of $\beta$ at the present iteration, so the natural step is to invert this via Bayes' rule;
\begin{equation}
    P\left(\beta \mid \log X_i\right) = \frac{\mathcal{P}\left(\log X_i \mid \beta\right) P\left(\beta\right)}{P\left(\log X\right)}.
\end{equation}
As with all Bayesian analyses, the distribution of $\beta$ is fixed up to a prior, which we choose to be uniform in $\beta$. The obtained temperatures are consistent with the previous two choices, which may seem oddly coincidental. In fact, closer inspection reveals that large values of $P(\beta \mid \log X_i)$ are the temperatures with a large value of the posterior at the present contour, normalised by the corresponding evidence. Thus the Bayesian temperature uses the same idea as the microcanonical one, except it accounts for the spread in the result.

@@

\subsection{Dimensionality}
We can immediately use the inferred temperature to track how the effective dimensionality of the posterior changes throughout the run, which was previously inaccessible. \citet{Handley_2019} demonstrated that at the end of a run, a measure of the number of constrained parameters is given by the Bayesian model dimensionality (BMD), defined as the posterior variance of the information content:
\begin{equation}
\frac{d_G}{2} = \int \mathcal{P}(\theta) \left(\log \frac{\mathcal{P}(\theta)}{\pi(\theta)} - \DKL\right)^2 \: \mathrm{d}\theta
= \langle \mathcal{I}^2 \rangle_\mathcal{P} - \langle \mathcal{I} \rangle^2_\mathcal{P}.
\end{equation}
\par
Calculating the quantity using intermediate set of weighted samples (which is concentrated at a single point) leads to vanishing variance, hence also dimensionality. However, we can recover the structure of the posterior together with the true dimensionality by adjusting the temperature. Dimensionality estimates are plotted in \cref{fig:d_G_spherical} for a spherical 32-d Gaussian, for which the true dimensionality is known. 
\par
The different choices of temperature are again consistent, but for the rest of this paper we choose the Bayesian $\beta$, because it provides a better reflection of the uncertainty in the estimate; the others, while fluctuating around the true value, are often many standard errors away from the true value at each single point.
\begin{figure}
\begin{center}
    \includegraphics{figures/d_G_spherical.pdf}
\end{center}
\caption{Dimensionality estimates using the different temperatures for a spherical 32-d Gaussian. Again, all are consistent, but the Bayesian definition has a uncertainty which includes the true value far more consistently than for the other definitions.}
\label{fig:d_G_spherical}
\end{figure}

>>>>

\section{Background}\label{sec:background}
We begin with a brief description of the nested sampling algorithm to establish the necessary notation. For a more comprehensive treatment, we recommend the original \citep{skilling} paper, the \citet{sivia} textbook, as well as the excellent technical review by \citet{Buchner_2023} and Nature review by \citet{physical_scientists}. 
\par
For a given likelihood $\mathcal{L}(\theta)$ and prior $\pi(\theta)$, nested sampling simultaneously calculates the Bayesian evidence
\begin{equation}\label{eq:evidence}
	\mathcal{Z} = \int \mathcal{L}\left(\theta\right)\pi(\theta)\ \mathrm{d}\theta
\end{equation}
while producing samples of the posterior distribution
\begin{equation}
	\mathcal{P}(\theta) = \frac{\mathcal{L}(\theta) \pi(\theta)}{\mathcal{Z}}.
\end{equation}
The algorithm operates by maintaining a set of $\nlive$ \textit{live points} sampled from the prior, which can vary in number throughout the run \citep{dynamic_ns}. At each iteration $i$, the point with the lowest likelihood is removed and added to a list of \textit{dead points} (illustrated in \cref{fig:dead_measure}). New points are then (optionally) drawn from the prior, subject to the constraint that they must have a higher likelihood than the latest dead point~$\mathcal{L}_i$. Repeating the procedure leads to the live points compressing around peaks in the likelihood. If there are more births than deaths, the number of live points $n_i$ increases, whilst choosing to not generate new points reduces the number of live points. The exact creation schedule depends on the dynamic nested sampling strategy.
\par
The integral in \eqref{eq:evidence} is then evaluated by transformation to a one-dimensional integral over the \textit{prior volume} $X$
\begin{equation}
	\mathcal{Z} = \int_0^1 \mathcal{L}(X)\ \mathrm{d}X \approx \sum_{i=1} \mathcal{L}_i\ \tfrac{1}{2}(X_{i-1}-X_{i+1}),
\end{equation}
where $X(\mathcal{L})$ is the fraction of the prior with a likelihood greater than $\mathcal{L}$. Whilst the likelihood contour $\mathcal{L}_i$ at each iteration is known, the prior volumes $X_i$ must be statistically estimated as follows: one can define a \textit{shrinkage factor} $t_i$ at each iteration $X_{i} = t_i X_{i-1}$, such that
\begin{equation}\label{eq:X_dist}
	X_i = \prod_{k=1}^i t_k.
\end{equation}
The $t_i$ are the maximum of $\nlive$ points drawn from $[0,1]$, so follow the distribution
\begin{equation}\label{eq:t_dist}
	P(t_i) = \nlive t_i^{\nlive-1},
\end{equation}
\begin{equation}\label{eq:t_moments}
    \langle\log t_i\rangle = -\frac{1}{\nlive}, \quad \mathrm{Var}(\log t_i) = \frac{1}{\nlive^2}.
\end{equation}
The algorithm terminates when an user-specified condition is met; a popular choice is when the evidence in the live points falls below some fraction $\epsilon$ of the accumulated evidence e.g. $10^{-3}$, which is proven to be a valid convergence criterion \citep{evans, Chopin_2010}.
Much of the existing literature treats this remaining evidence separately, for instance by estimating it as the termination $X$ multiplied by the average likelihood amongst the remaining live points. It is, however, quantitatively equivalent but qualitatively neater to consider termination as killing the remaining live points off one-by-one, incrementing the evidence exactly as during the run with decreasing $\nlive$ \citep{dynesty}.
\par
Uncertainties in the evidence are dominated by the spread in the prior volume distribution, and the simplest way to estimate them is by Monte Carlo sampling over sets of $\bm{t}$. As Skilling and others \citep{Chopin_2010, Keeton_2011} have shown, for any given problem the uncertainty in $\log \mathcal{Z}$ is proportional to $1/\sqrt{n_\mathrm{live}}$, so $n_\mathrm{live}$ sets the resolution of the algorithm.

\section{The anatomy of a nested sampling run}\label{sec:anatomy}
The following sections act as an inventory of the information available to us at an intermediate iteration $i$, which we shall use to make endpoint predictions in \cref{sec:endpoint}. We present an anatomy of the progression of a nested sampling run in terms of the prior volume compression (\cref{sec:prior_volume}), the log-likelihood increase (\cref{sec:logL}), the inferred temperature (\cref{sec:temperature}), and the dimensionality of the samples (\cref{sec:dimensionality}).

\subsection{Prior volume}\label{sec:prior_volume}
The key feature of nested sampling is that the sampling is controlled by prior volume compression. The task is to find the posterior typically lying in a tiny fraction of the prior volume, a total compression which is quantified by the average information gain, or \textit{Kullback-Leibler divergence}:
\begin{equation}\label{eq:DKL}
   \DKL = \int \mathcal{P}(\theta) \log \frac{\mathcal{P}(\theta)}{\pi(\theta)}\ \mathrm{d}\theta. 
\end{equation}
The bulk of the posterior lies within a prior volume ${X = e^{-\DKL}}$, which is the target compression. From \cref{eq:t_moments} one gets there by iteratively taking steps of size ${\Delta \log X_i = -1/n_i}$, so that when we add up the contribution of each step in \eqref{eq:t_moments} we get
\begin{equation}
    \langle\log X_i\rangle = -\sum_{k=1}^i \frac{1}{n_k}, \quad \mathrm{Var}(\log X_i) = \sum_{k=1}^i \frac{1}{n_k^2}.
\end{equation}
A constant step size in $\log X$ corresponds to a geometrically constant measure for the dead points , which is exactly needed to overcome the curse of dimensionality.
\begin{figure*}
\begin{center}
    \includegraphics{figures/dead_measure_live.pdf}
\end{center}
\caption{The dead points of a nested sampling run, recursively zoomed in. Their density is constant in $\log X$, which is a geometrically constant measure $\propto \mathrm{d} X/X$, and hence scale invariant until the final set of live points is reached. The live points at a given iteration $i$ (larger dots) are uniform across the prior, plotted for comparison.}
\label{fig:dead_measure}
\end{figure*}
\par
The same is not true for the live points, which are uniformly distributed in prior volume, and by the comments in the penultimate paragraph of \cref{sec:background} have ${n_{i+k} = n_i-k}$. As a result, the maximum live point is found at 
\begin{equation}\label{eq:Xmin}
    \langle\log X_\mathrm{min}^{\mathrm{live}}\rangle = \langle\log X_i\rangle - \sum_{k=1}^{\nlive} \frac{1}{n_{i+k}} \approx -\frac{i}{\nlive} - \log \nlive - \gamma,
\end{equation}
with variance
\begin{equation}
    \mathrm{Var}(\log X_\mathrm{min}^{\mathrm{live}}) = \mathrm{Var}(\log X_i) + \sum_{k=1}^{\nlive} \frac{1}{n_{i+k}^2} \approx \frac{i}{\nlive^2} + \frac{\pi^2}{6}, 
\end{equation}
where the large $\nlive$ limit is taken for the approximation to the harmonic series, $\gamma$ being the Euler-Mascheroni constant.
\par
The live points therefore only get us a factor of $\log \nlive$ closer in volume to the posterior bulk. In other words, it is not until we are around $\log \nlive$ away from $\log X = -\DKL$ that the samples begin populating the posterior typical set. One can see from \eqref{eq:DKL} that the divergence increases linearly with dimension, so for large dimensionalities and typical live point numbers $\lesssim 1000$, this does not happen until near the end of the run.
\par
The result is consistent with that in \citet{statmech}, which states that a spike at a volume smaller than $X_i/n_i$ will go undetected. Intuitively, it is because for a sharply peaked likelihood the live points are too diffuse to land there with any significant probability for most of the run. These results are summarised in \cref{fig:logX_distribution}
\begin{figure*}
\begin{center}
    \includegraphics{figures/logX_distribution.pdf}
\end{center}
\caption{The distribution of the posterior mass in terms of $\log X$, the live points over the constrained prior and the smallest live point prior volume $\log X_\mathrm{min}^{\mathrm{live}}$ at an intermediate iteration $i$. For large values of $\DKL$ i.e. informative posteriors and/or large dimensionalities, the maximum live point is very far from the posterior bulk until the very end of the run. Note that the x-axis in this plot is $\log X$, so that the run proceeds from right to left to emphasise that the enclosed prior volume iteratively gets smaller. Plots of the sort from here onward will be in terms of $-\log X$, where the run will more naturally proceed from left to right. }
\label{fig:logX_distribution}
\end{figure*}

\subsection{Log-likelihood}\label{sec:logL}
We now consider the distribution of the live and dead points in log-likelihood. We will examine the representative case of the $d$-dimensional multivariate Gaussian:
\begin{equation}\label{eq:gaussian_logL}
    \log \Like = \logLmax - X^{2/d}/2\sigma^2
\end{equation}
with maximum point $\logLmax$ and length scale $\sigma$ to get an insight into the analytics. The likelihood monotonically increases towards $\logLmax$, with \cref{sec:logLmax} showing the posterior samples eventually concentrated around the bulk at
\begin{equation}
    \langle\log\Like\rangle_\mathcal{P} = \log\mathcal{L}_\mathrm{max} - \frac{d}{2},  \quad \mathrm{Var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}.
\end{equation}
We can again find the size of each step in log-likelihood, as well as the expected location of the maximum live point. Let us define a likelihood normalised by the distance to the maximum:
\begin{equation}
    y = \frac{\log\mathcal{L}-\log\mathcal{L}_i}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_i};
    \label{eq:normalised_likelihood}
\end{equation}
$y=0$ corresponds to the current point $\log\mathcal{L}_i$ and $y=1$ to the maximum $\log\mathcal{L}_\mathrm{max}$. At each iteration, $y$ increases by roughly
\begin{equation}
    \lim_{\nlive \to \infty}\Delta y \approx \frac{\mathrm{d} y}{\mathrm{d}\log X} \Delta \log X = \frac{2}{d \nlive}  
\end{equation}
in the large $\nlive$ limit. We can again sum the harmonic series to see that the maximum live point is expected to be at 
\begin{equation}
    y_{(\nlive)}^{\mathrm{live}} = \frac{2\log \nlive}{d},
\end{equation}
where we have used the notation of order statistics to denote $x_{(n)}$ as the maximum of $n$ points. This is a very small fraction in high dimensions, showing again that until the end the live points are far from the posterior bulk, and certainly nowhere near the maximum.
\par
The above also implies that the normalised distance between the highest and second highest live point is roughly 
\begin{equation}
    y_{(\nlive)}^{\mathrm{live}} - y_{(\nlive - 1)}^{\mathrm{live}} \approx \frac{2}{d}.
\end{equation}
Before reaching the posterior bulk, $\logLmax - \log\Like_i > d/2$, so we must have
\begin{equation}
    \log \Like^{\mathrm{live}}_{(\nlive)} - \log \Like^\mathrm{live}_{(\nlive - 1)} > 1.
\end{equation}
In other words, the highest likelihood point is always at least an order of magnitude greater than the second highest. It is therefore typically the case that nearly all of the posterior mass is concentrated in a single point, the maximum live point, until the very end of the run when the prior volumes have shrunk enough to compensate.

@@

\subsection{Temperature}\label{sec:temperature}
\subsubsection*{Motivations}
As shown in the previous section, midway through the run nearly all of the posterior mass is concentrated at a single point. However, this does not capture the \textit{structure} of the posterior that has been explored and all of the information it provides. 
\par
We have the potential to fix this because nested sampling is invariant to monotonic transformations, so we can transform the likelihood as $\Like \to \Like^{\beta}$ without loss of information by trivially re-weighting the samples. Increasing $\beta$ worsens the situation, while $\beta \to 0$ simply gives back the prior. There is, on the other hand, a significant intermediate range which makes the samples look like a posterior centred at the present contour, which will allow us to recover the structure of the samples. A schematic of the procedure is shown in \cref{fig:last_live_point}.
\par
\begin{figure}
\begin{center}
    \includegraphics{figures/last_live_point.pdf}
\end{center}
\caption{The likelihood and posterior as a function of $\log X$ in the middle of a nested sampling run. Almost all of the posterior mass is concentrated at a single point with the highest compression, because it is orders of magnitude higher in likelihood. Reducing $\beta$ re-weights to shift the posterior mass; taking $\beta \to 0$ goes too far and gives back the prior, but there is some intermediate $\beta = \beta_i$ which is significant because it centres the posterior at the current contour.}
\label{fig:last_live_point}
\end{figure}
\par
At this point it is relevant to note the correspondence between Bayesian inference and statistical mechanics, from which the above transform is derived. If one equates the parameters to microstates $i$, the negative log-likelihood to the microstate energy $E_i$, and the prior to the density of states $g_i$, then the posterior as given by the generalised Bayes' rule is the canonical ensemble
\begin{equation}
    p(E_i) = \frac{g_i e^{-\beta E_i}}{Z(\beta)} \quad \leftrightarrow \quad \mathcal{P}_\beta(\theta) = \frac{\mathcal{L}^{\beta}(\theta)\pi(\theta)}{\mathcal{Z}(\beta)}
\end{equation}
at the inverse temperature $\beta = 1/T$. As noted by \citet{demons}, thermal algorithms such as thermodynamic integration \citep{path_sampling} get the evidence by evolving through a series of canonical ensembles via some temperature schedule, but nested sampling instead maintains \textit{microcanonical} ensembles, which are the iso-likelihood contours. Instead of using temperature \citep{simulated_annealing, parallel_tempering} or energy \citep{wang_landau} as a control parameter, nested sampling chooses a series of ensembles with constant relative volume entropy $\Delta \log X$, which allows the algorithm to handle phase transitions \citep{baldock}.
\par
Because the temperature of a microcanonical ensemble is a derived property rather than a parameter, there is some freedom in its definition. Returning to our original motivation, we make the connection that the temperature is the re-weighting $\Like \to \Like^{\beta}$ which centres the ensemble around the current energy. We now present several temperatures that achieve this aim, each of which one can plausibly consider to be the current temperature of a nested sampling run.

\subsubsection{Microcanonical temperature}\label{sec:microcanonical_temperature}
The obvious candidate is the microcanonical temperature $\partial S/\partial E$, where the volume entropy is $\log X$ and the energy is as usual $-\log \Like$. This gives the density of states; as discussed in Skilling's original paper,
\begin{equation}
    \beta_\mathrm{M}  = - \frac{\mathrm{d} \log X}{\mathrm{d} \log \Like} \Bigg\vert_{\log \Like_i}.
\end{equation}
is the $\beta$ at which $\Like^{\beta} X$ peaks at $\log X_i$, if we assume differentiability, which is exactly the intuition we were aiming for to put the ensemble bulk at the current contour. 
\par
Its value can be easily obtained via finite difference of the $\log \Like$ and $\log X$ intervals, albeit subject to an arbitrary window size for the differencing. Indeed, material science applications \citep{Baldock_2017} use this estimator to monitor the `cooling' progress of nested sampling, with a window size of 1000 iterations.

\subsubsection{Canonical temperature}\label{sec:canonical_temperature}
Another temperature considered by \citet{demons} is that at which the current energy (i.e. $-\log \Like_i$) is the average energy of the entire ensemble. One can obtain it by inverting 
\begin{equation}
    \langle \log \Like \rangle_{\mathcal{P}_\beta} = \log \Like_i
\end{equation}
to get the `canonical' temperature $\beta_\mathrm{C}$. While $\beta_\mathrm{M}$ is derived from (the gradient of) a single contour, this temperature uses the entire ensemble. It has the desirable property that it rises monotonically with compression, in analogy to a monotonic annealing schedule.

\subsubsection{Bayesian temperature}\label{sec:bayesian_temperature}
We furthermore propose a temperature $\beta_\mathrm{B}$ that is obtained via Bayesian inference, which returns a distribution rather than a point estimate. Since each value of $\beta$ leads to a different likelihood $\Like^{\beta}$, one can consider the posterior distribution as a function of $\log X$ to be \textit{conditioned} on $\beta$. We can therefore write
\begin{equation}
   \mathcal{P}(\log X \mid \beta) = \frac{\Like^{\beta}(X) X}{\mathcal{Z(\beta)}}.
\end{equation}
What we would really like is the distribution of $\beta$ at the present iteration, so the natural step is to invert this via Bayes' rule;
\begin{equation}
    P\left(\beta \mid \log X_i\right) = \frac{\mathcal{P}\left(\log X_i \mid \beta\right) P\left(\beta\right)}{P\left(\log X\right)}.
\end{equation}
As with all Bayesian analyses, the distribution of $\beta$ is fixed up to a prior, which we choose to be uniform in $\beta$. The obtained temperatures are consistent with the previous two choices, which may seem oddly coincidental. In fact, closer inspection reveals that large values of $P(\beta \mid \log X_i)$ are the temperatures with a large value of the posterior at the present contour, normalised by the corresponding evidence. Thus the Bayesian temperature uses the same idea as the microcanonical one, except it accounts for the spread in the result.

@@

\subsection{Dimensionality}\label{sec:dimensionality}
We can immediately use the inferred temperature to track how the effective dimensionality of the posterior changes throughout the run, which was previously inaccessible. \citet{Handley_2019} demonstrated that at the end of a run, a measure of the number of constrained parameters is given by the Bayesian model dimensionality (BMD), defined as the posterior variance of the information content:
\begin{equation}
\frac{d_G}{2} = \int \mathcal{P}(\theta) \left(\log \frac{\mathcal{P}(\theta)}{\pi(\theta)} - \DKL\right)^2 \: \mathrm{d}\theta
= \langle \mathcal{I}^2 \rangle_\mathcal{P} - \langle \mathcal{I} \rangle^2_\mathcal{P}.
\end{equation}
\par
Calculating the quantity using intermediate set of weighted samples (which is concentrated at a single point) leads to vanishing variance, hence also dimensionality. However, we can recover the structure of the posterior together with the true dimensionality by adjusting the temperature. Dimensionality estimates are plotted in \cref{fig:d_G_spherical} for a spherical 32-d Gaussian, for which the true dimensionality is known. 
\par
The different choices of temperature are again consistent, but for the rest of this paper we choose the Bayesian $\beta$, because it provides a better reflection of the uncertainty in the estimate; the others, while fluctuating around the true value, are often many standard errors away from the true value at each single point.
\begin{figure}
\begin{center}
    \includegraphics{figures/d_G_spherical.pdf}
\end{center}
\caption{Dimensionality estimates using the different temperatures for a spherical 32-d Gaussian. Again, all are consistent, but the Bayesian definition has a uncertainty which includes the true value far more consistently than for the other definitions.}
\label{fig:d_G_spherical}
\end{figure}
